Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions

### 2022-06-03 analysis/paper/shceof.Rmd
Agerga data availabilty y modifica goyal

<!-- The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts and could even be related to the occurrence of high-impact climate extremes (REFERENCE).  -->

<!-- Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980]. -->

<!-- Hablar simplemente de cómo estos fenómenos varían en longitud.  -->

<!-- @hill2009 ->  -->

<!-- Comparan eventos ENSO del 97 y de 2002 -->

<!-- La posición de las anomalías de circulación (PSA) son importantes para determinar el impacto del ENSO en sudamérica. Sugiere que la localización de la anomalía de SST cambia la localización del PSA.  -->

<!-- > The meridional winds from the idealised experiments based on simplified and localised SST warming showed a significant difference between experiments forced by eastern versus central-west Pacific SST anomalies, with a clear phase shift of the PSA apparent. This phase shift in the idealised experiments is possibly the main factor responsible for the different responses in South American atmospheric circulation between the 1997/1998 and 2002/2003 El Niño events. The phase shift over southeastern South America was also seen in the low-level 850 hPa winds in the EN97 and EN02 experiments, suggesting that the PSA phase shift is responsible for the shift in the moisture transport over southeastern Brazil between the two El Niño events. In summary, the location of the maximum SST anomaly appears to be an important factor in controlling the amount of moisture transported into the SACZ region during El Niño events.  -->

<!-- @hobbs2010 -> muestra la importancia de la longitud de ciclones -->

<!-- ciasto2015 -> no investiga impactos -->

<!-- sun2013 -->

<!-- wilson2014 -> -->

<!-- Observaciones y modelos enso canonico y modoki. diferentes patrones de circulacion.  -->

<!-- wang2021 ->  -->

<!-- > However, the spatial pattern of the PSA wave train excited by ENSO seems to be -->

<!-- > non-stationary. Fogt and Bromwich (2006) indicated that the PSA mode moved -->

<!-- > eastwards by 20° longitude in the 1990s owing to the eastward shift of the tropical -->

<!-- > convective position. In addition, Wilson et al. (2014, 2016) and Ciasto et al. (2015) -->

<!-- > found that the different types of ENSO can contribute to the wave shift. Specifically, -->

<!-- > Rossby waves excited by central Pacific (CP)–type El Niño shift westwards owing to -->

<!-- > the westward movement of the tropical diabatic heating. -->

<!-- Diferencia en la localización de los centros en LN y EN -> diferencias en los impactos en la península antártica.  -->

<!-- Lo más importante es la localización de la salida del jet, no la perturbación de sst.  -->

<!-- fogt2006 ->  -->

<!-- cambios en la posición de las teleccionexiones. Pero es todo con composiciones y comparando décadas. No mira impactos. -->

<!-- lou2021 ->  -->

<!-- > In combination, they produce the single propagating PSA mode. -->

<!-- Complicado y no entiendo mucho, pero cosas que puedo ver es que PSA forza el océano subtropical que luego es precursora de ENSO. No sé si eso es súper  -->

@@

Our method is similar to @goyal2022 as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2. 
Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.

@@

>>>>

@@

The method used in this study have similarities to the one used by @goyal2022 as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies. 
The patterns obtained by them bear high resemblance with cEOF2. 
Although a detailed comparison is out of scope for this paper, the cEOF analysis has the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.

@@

## Availability of data and material {.unnumbered}

All data used in this paper available in a version-controlled repository [@campitelli2022a]. 
It is also freely available from their respective sources:

- ERA5 data can be obtained via the Copernicus Climate Data Store (<https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\>).

- ERSSTv5 can be obtained via NOAA's NCEI websiste at <https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00927>

- CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
